---
title: "[Method Name] for Nutrition Research"
author: "[Author Name]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center"
)

# Shared random seed — set this once and pass it wherever randomness is needed
SEED <- 42
set.seed(SEED)
```

## Introduction

<!--
~200 words. Two paragraphs:
1. What is [Method] and when would a nutrition researcher use it?
2. What specific question are we answering with NHANES data in this tutorial?
No subheadings in this section.
-->

[This tutorial demonstrates **[Method Name]** using data from the National Health
and Nutrition Examination Survey (NHANES). We will predict [outcome] using
[predictors], walking through each step from data loading to interpretation.]

---

## Setup

```{r packages}
suppressPackageStartupMessages({
  library(tidyverse)   # Data manipulation and visualization
  library(here)        # Reproducible file paths
  # ADD METHOD-SPECIFIC PACKAGES BELOW, each with a comment:
  # library(randomForest)  # Random forest models
  # library(caret)         # Cross-validation and model evaluation
})
```

---

## Data

```{r load-data}
# Load NHANES data — tries GitHub URL first, falls back to local copy
DATA_URL <- "https://raw.githubusercontent.com/m-powell/m-powell.github.io/master/MLforNutrition/data/nhanes_nutrition.csv"

nhanes <- tryCatch(
  read_csv(DATA_URL, show_col_types = FALSE),
  error = function(e) {
    message("Could not load from URL; falling back to local copy.")
    read_csv(here::here("data", "nhanes_nutrition.csv"), show_col_types = FALSE)
  }
)

# Quick inspection
glimpse(nhanes)
```

<!--
Briefly describe the dataset: relevant columns, time period, sample size.
Describe any preprocessing steps BEFORE the code that performs them.
-->

```{r preprocess}
# [Describe what preprocessing is happening and why before this chunk]

nhanes_clean <- nhanes |>
  drop_na(bp_sys, bp_di, age, bmi, kcal) |>
  mutate(
    extreme_bp = factor(extreme_bp, levels = c(FALSE, TRUE), labels = c("Normal", "Extreme"))
  )

cat(sprintf("Observations after cleaning: %d\n", nrow(nhanes_clean)))
```

---

## Method

<!--
Break this section into level-3 subheadings. Suggested structure:
  ### Model Setup and Hyperparameters
  ### Fitting the Model
  ### Evaluating Performance
  ### Interpreting Results

Before every code chunk, write a sentence or two explaining what you're doing and why.
-->

### Model Setup and Hyperparameters

<!-- Explain the key hyperparameters conceptually before setting them. -->

```{r model-setup}
# [Hyperparameter setup code here]
```

### Fitting the Model

```{r fit-model}
set.seed(SEED)  # Always set seed immediately before model fitting

# [Model fitting code here]
```

### Evaluating Performance

```{r evaluate}
# [Evaluation code here]
```

### Interpreting Results

```{r visualize}
# ---- Shared plot theme (use for every figure in this tutorial) ----
theme_nutr <- function() {
  theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(size = 14, face = "bold"),
      plot.subtitle    = element_text(size = 12, color = "gray40"),
      plot.caption     = element_text(size = 10, color = "gray50", hjust = 0),
      axis.title       = element_text(size = 12),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
}
# -------------------------------------------------------------------

# [Visualization code here]
# Caption format: "Figure N: [What is shown]. [One sentence interpretation]."
```

---

## Results

<!--
1–2 paragraphs. Summarize what the model found in plain language.
Keep it grounded in the nutrition/health context.
-->

---

## Summary

**Key takeaways:**

- ...
- ...
- ...

**Limitation:** [One sentence on when this method is not appropriate.]

**See also:** The [Python companion tutorial](../python/[method]-python.html)
covers the same analysis using scikit-learn.

---

```{r session-info, echo=FALSE}
sessionInfo()
```
